Creating Exploratory Graphs
========================================================

Why do we need exploratory graphs in the first place?
* to better understand the data for ourselves
* to find patterns in data
* to intuite modelling strategies as a next step
* to help debug analyses
* to communicate results

## Principles of Analytical graphics

1 Show comparisons (always ask "relative to what")
2 Lead with an explanation or Hypothesis of what is going on
3 Show multi-variate data. Life is far more complicated that bivariate!
4 Integrate evidence across multiple modes (words, diagrams, tables,etc)
5 Describe and document evidence with appropriate data and labels
6 Content is King!!!

## Types of graphics

Now, what are some of the ways in which we can start to explore a dataset visually or otherwise? Let's see:
* Summary in R: 1st, 2nd (median), 3rd quantiles, min-max and Average.
* Histogram:  how are values distributed, what's a good pdf for pop ?
* Scatter plots: good indication of relation between X&Y
* Boxplot: plots 25-75 percentile with line for median. Whiskers for outliers
* Barplots: 

**Action**: Need to get a better sense of other types and their specific applicability in EDA.


## Types of plotting systems in R

We got 3 major systems that we explore in this document. 

1 **Base** plotting system  
2 **Lattice** plotting system  
3 **ggplot2** system  

Let's review them at a high level below.

## Base plotting system

This system is best thought of as an artist's palette. This system requires a two step process to create a plot: where the first step is to create the basic plot and the next step is to annotate it (through legenda and formatting). 

Pros:
* Lets us have pretty granular control of the plot
* we can have unrelated panels and lots of them

Cons:
* does take some coding
* does not provide  defaults (restatement of point 1)

## Lattice plotting system

This is a good system for reviewing plots of X,Y by values of a conditioning variable Z (where z is a factor with levels).

Pros:
* Everything is done in ONE function call 
* Defaults and calculates lots of the formatting option on its own
* Is good for looking at variation of data by levels of conditioning variables

Cons:
* one line function calls are pretty cryptic
* flexibility is limited

## ggplot2 

This package implements "grammar of graphics", which  is a construct that has a theory of how graphics need to be built. Defines concepts and relations between concepts. Sort of mid-way between the Base and Lattice systems.

Pros:
* Intuitive, shorterst path from mind to page!
* great defaults but allows pretty fine-grained customization as well
* In my opinion, looks pretty cool!


## Base plotting system in action

### Examples

Let us get into it here with a basic scatter plot from our airquality dataset:

```{r baseplots,cache=TRUE}
data(airquality)
par(mfrow=c(1,1)) # single pane plot
# create a base plot but suppress output
with(airquality,plot(Wind,Ozone,main="Ozone and Wind in NY city",type="n"))
# add points based on conditions > here month of May Vs all other 
with(subset(airquality,Month==5),points(Wind,Ozone,col="blue", pch=2))
with(subset(airquality,Month!=5),points(Wind,Ozone,col="red",pch=3))
# add legend
legend("topright",col=c("blue","red"),pch=c(2,3) ,legend=c("May","other months"))

# Fit a linear regression line
model<-lm(Ozone~Wind,airquality)
# add line to the plot
abline(model,lwd=2)
```


Let us review a few more type of plots very quickly:

Scatter plot using the "plot" function:
```{r base_scatterplot, cache=TRUE}
par(mfrow=c(1,2)) # setting up a two panel plot
with(airquality,{
        
        plot(Wind,Ozone,main="Ozone and Wind")
        plot(Solar.R,Ozone,main="Ozone and Solar Radiation")
        
})

```


Histogram using the  "hist" function:
```{r base_histogram,cache=TRUE}
par(mfrow=c(1,1)) # Single plot
hist(airquality$Ozone)

```

Boxplot using the  "boxplot" function:
```{r base_boxplot,cache=TRUE}
par(mfrow=c(1,1)) # Single plot
testdf<-transform(airquality,Month=factor(Month))
boxplot(Ozone~Month,testdf)
```

### Graphics parameters

There are some basic graphics parameters that are used in this plotting system. A partial list  follows:  

* pch: is the plotting symbol
* col: refers to the color attribute
* lty: line type
* xlab.yab: X and Y labels
* mar: margins
* mfrow,mfcol: defines the number of panels row-wise vs col wise
* par("<parameter>"): provides current values of parameter
* par(parameter=new value) sets new values for parameter

### Graphics Devices

The base plotting package prints the plots to "graphic devices". These can either be the screen device or a file device. 

* Screen Device: This is called quartz() in mac, windows() in windows etc
* File devices: These are further broken down into Vector and Bitmap devices
* Vector File devices: line type interfaces, Resizes well, Portable
* Bitmap File devices: Good for line drawings and images with solid color. Does not resize well

We can use dev.copy() to copy from screen device to a file device. Sometimes, when we do this, the plot does not appear exactly. In that case we directly call the file device to plot on it. 

## Lattice plotting system

This system is good for plotting several plots at the same time by levels of a conditioning variable. As an example:  
xyplot(y~x|f.g,data=dataframe) plots the scatter of y Vs x across all levels of f.g  

```{r latticescatter,cache=TRUE}

library("lattice")
data(airquality)
xyplot(Ozone~Wind,data=airquality) # basic scatter plot
testdf<-airquality
testdf<-transform(testdf,Month=factor(Month))

# scatter plot by month
xyplot(Ozone~Wind|Month,data=testdf,layout=c(5,1))



```


### Some key functions in lattice package

* xyplot: for scatterplot per our example above
* histogram
* bwplot: box and whiskers plot
* Others like stripplot, dotplot, splom, levelplot, contour plot

### Additional notes on lattice package

* The functions above return object of type "trellis"
* these objects are autoprinted in command line and hence the plot appears on screen
* the trellis objects can technically be assigned on variables and stored for later use
* print functions within the lattice packages get the plot on to the various graphics devices
* panel in lattice functions allow us to write custom functions that help control appearance of the  panels. there are default implementations that are involved if we dont override.  

## ggplot2 package

This package implements the "grammar of graphics" which is an abstraction of graphical concepts, ideas and their inter-relations.  
  
Statistical graphs are "mapping of data to aesthetic attributes ( color, shape) of geometric objects (points, lines, bars etc)".  
Plots may also contain statistical transformation of data and is drawn on a coordinate system.    

* the qplot() is the quickplot function that is easy, good-looking (+ lots of defaults pre-set)
* ggplot() is the work-horse function, that can be used to customize

### The qplot() function

* Data supplied in dataframes
* Made up of aesthetics (size, shape, colors) and geoms(points, lines)
* Facets parameters use the Factor levels of the conditioning variables and are similar in concept to what we have in lattice > supports f.g operation

```{r ggplot2intro,cache=TRUE}

library(ggplot2)
qplot(Wind,Ozone,data=testdf,facets=.~Month)
# testdf has been defined before with months as factors

```

The figure above is a very clean looking output which is why I prefer the ggplot2 function. We can add smoothers to the plots as below.  

```{r ggplot2smoother,cache=TRUE}

qplot(Wind,Ozone,data=testdf,facets=.~Month, geom=c("point","smooth"))
# the default smoother is "loess"

```

we can add linear smoothers as well as follows.  

```{r ggplot2linsmoother,cache=TRUE}

qplot(Wind,Ozone,data=testdf,facets=.~Month, geom=c("point","smooth"), method="lm")
# changing the smoothing method to  "linear model"

```

The colors of the geoms can also be modified by assigning them to a conditioning variable as follows. We demonstrate this with our iris data set that has three species of flowers represented in the 150 sample dataframe.

```{r ggplot2facetcolora,cache=TRUE}

data(iris) # loading iris dataset
qplot(Sepal.Length,Sepal.Width, data=iris,col=Species, geom=c("point","smooth"))
# colors are based on values of species
# also added a default loess smoother

```

Histograms are created by referencing just one variable in the qplot function.

```{r ggplot2histogram,cache=TRUE}

qplot(Sepal.Length,data=iris,facets=.~Species)
# histogram is based on length of Sepals

```

We can also have a joint histogram with conditioning variable highlighted with the larger distribution as below:

```{r ggplot2histogramconditioning,cache=TRUE}

qplot(Sepal.Length,data=iris,fill=Species)
# the histogram color is controlled by parameter fill which varies based on species

```

With qplot, we can also plot density functions of the variable :

```{r ggplot2density,cache=TRUE}

qplot(Sepal.Length,data=iris,geom="density")

```

which we can also condition as below:

```{r ggplot2densitycondition,cache=TRUE}

qplot(Sepal.Length,data=iris,geom="density", col=Species)

```
